Impact of Molecular Subtype on Inflammatory Breast Cancer-Specific Mortality in Adult Women

### 2024-12-03 products/manuscript/manuscript.qmd
manuscript edits

---
title: "epid 8010 manuscript"
author: " "
format: pdf
bibliography:
  - "refs.bib"
---

@@

**Author**

Murphy John$^{1}$

**Author affiliation**

1.  Department of Epidemiology and Biostatistics, University of Georgia, Athens, GA, USA.

@@

Therefore, using the most recent Surveillance, Epidemiology, and End Results (SEER) data, this study aims to assess the effect of IBC molecular subtype on breast cancer-specific mortality in adult females with IBC.

@@

@@

A total of 5969 adult female IBC cases were eligible during 2010-2021 in the SEER database for our study population. Of these, 1829 were excluded; 165 had a survival time less than 1 month, 487 had an unknown or competing cause of death, and 1177 had incomplete information. The remaining 4140 were included in our analysis, as shown in Figure 1.

```{r}
#| label: fig-flowchart
#| fig-cap: |
#|   [Flowchart of IBC patients screening on SEER database]{color="#FF2400"}
#| out.width: 100%
#| out.height: 80%
knitr::include_graphics(
	here::here("results", "figures", "flowchart.png")
)
```

The clinicopathological characteristics of eligible patients are summarized in Table 1. The age subgroup of 20–49 years accounts for 35.1% of Luminal B cases, a higher proportion compared to other subtypes. Moreover, Luminal B and HER2-positive subtypes appear to affect younger age groups more frequently than Luminal A, with 85.4% and 84.5% of cases, respectively, occurring in patients under 69 years, compared to 76.5% for Luminal A. A higher proportion of Black patients is observed in the Triple-Negative group compared to other subtypes. Marital status analysis shows that the Triple-Negative group has the highest percentage of unmarried women. Triple-Negative cases also have the highest percentage of N3 node stage and exhibit the greatest proportion of grade III/IV cases.

```{r}
#| label: tbl-table1
#| tbl-cap: |
#|   [Clinicopathological characteristics by cancer subtype]{color="#FF2400"}
knitr::include_graphics(
	here::here("results", "tables", "t1.png")
)
```

@@

The proportional hazards assumption was assessed for each covariate visually using log minus log plots, as shown in Figure 2.

```{r}
#| label: fig-flowchart
#| fig-cap: |
#|   [Log-log plots]{color="#FF2400"}
#| out.width: 100%
#| out.height: 80%
knitr::include_graphics(
	here::here("results", "figures", "loglog_plot.png")
)
```

Upon inspection, slight deviations from parallelism were observed for the covariates subtype, radiation, chemotherapy, and surgery. Therefore, the full Cox proportional hazards model was fitted to the data, and Schoenfeld residuals were extracted.  The correlation between Schoenfeld residuals and ranked survival times for each covariate is presented in Table 2.

@@

Table 2 presents the median survival in years, along with 1-, 5-, and 10-year survival percentages, stratified by clinicopathological characteristics, all of which show statistically significant differences based on the log-rank test. Survival outcomes differ significantly by subtype, with the Triple-Negative subtype showing the poorest prognosis, including a median survival of 1.7 years and 1-, 5-, and 10-year survival rates of 67%, 24%, and 21%, respectively. Conversely, the Luminal B subtype has the most favorable outcomes, with a median survival of 9.4 years and a 10-year survival rate of 50%. Racial disparities are evident, as Black patients have the poorest survival outcomes, with a median survival of 2.4 years compared to 4.6 years for White patients and 5.4 years for patients of other races. Unmarried women experience poorer survival compared to married women. Although 1-year survival is similar across all node stages, nodal involvement significantly affects long-term outcomes, with the N3 stage showing the lowest median survival of 2.7 years and the poorest 5- and 10-year survival rates. Survival differences are also pronounced by distant metastasis (M stage), with M1 stage associated with worse outcomes. Additionally, higher cancer grades correlate with poorer survival. In terms of treatment, patients who receive radiation, chemotherapy, or surgery exhibit better survival outcomes across all measurements.

```{r}
#| label: tbl-table2
#| tbl-cap: |
#|   [Survival for IBC by Clinicopathological Characteristics]{color="#FF2400"}
knitr::include_graphics(
	here::here("results", "tables", "t2.png")
)
```

We conducted univariate and multivariate analyses based on the data in Table 2, with the results of the Cox proportional hazards regression presented in Table 3.

```{r}
#| label: tbl-table3
#| tbl-cap: |
#|   [Cox Proportional Hazards Regression
#|   Model for Survival in Adult Women with IBC]{color="#FF2400"}
knitr::include_graphics(
	here::here("results", "tables", "t3.png")
)
```

@@

In this large population-based cohort of women diagnosed with IBC, we found a better survival in Luminal B molecular subtype and poorer in Triple Negative. We found that Found that Black patients made up the smallest percentage of Luminal B subtype and the greatest of Triple Negative. Similar findings were observed for patients who were unmarried at diagnosis.

It would be interesting to look at the combination of treatments (radiation, chemo, surgery)

>>>>

---
title: "Impact of Molecular Subtype on Inflammatory Breast Cancer-Specific Mortality in Adult Women"
author: "Murphy John"
format: pdf
bibliography:
  - "refs.bib"
---

@@

@@

Therefore, using the most recent Surveillance, Epidemiology, and End Results (SEER) data, this study aims to assess the effect of molecular subtype on IBC-specific mortality in adult females with IBC.

@@

\clearpage

@@

A total of 5969 adult female IBC cases were eligible during 2010-2021 in the SEER database for our study population. Of these, 1829 were excluded; 165 had a survival time less than 1 month, 487 had an unknown or competing cause of death, and 1177 had incomplete information. The remaining 4140 were included in our analysis, as shown in @fig-flowchart.

```{r}
#| label: fig-flowchart
#| fig-cap: |
#|   [Flowchart of IBC patients screening on SEER database]{color="#FF2400"}
#| out.width: 90%
#| out.height: 70%
knitr::include_graphics(
	here::here("results", "figures", "flowchart.png")
)
```

The clinicopathological characteristics of eligible patients are summarized in @tbl-table1. The age subgroup of 20–49 years accounts for 35.1% of Luminal B cases, a higher proportion compared to other subtypes. Moreover, Luminal B and HER2-positive subtypes appear to affect younger age groups more frequently than Luminal A, with 85.4% and 84.5% of cases, respectively, occurring in patients under 69 years, compared to 76.5% for Luminal A. A higher proportion of Black patients is observed in the Triple-Negative group compared to other subtypes. Marital status analysis shows that the Triple-Negative group has the highest percentage of unmarried women. Triple-Negative cases also have the highest percentage of N3 node stage and exhibit the greatest proportion of grade III/IV cases.

```{r}
#| label: tbl-table1
#| tbl-cap: |
#|   [Clinicopathological characteristics by cancer subtype]{color="#FF2400"}
#| echo: false

knitr::include_graphics(
	here::here("results", "tables", "t1.rds")
)
```

\clearpage

@@

The proportional hazards assumption was assessed for each covariate visually using log minus log plots, as shown in @fig-loglog.

```{r}
#| label: fig-loglog
#| fig-cap: |
#|   [Log minus log plots]{color="#FF2400"}
#| out.width: 100%
knitr::include_graphics(
	here::here("results", "figures", "loglog_plot.png")
)
```

Upon inspection, slight deviations from parallelism were observed for the covariates subtype, grade radiation, chemotherapy, and surgery. Therefore, the full Cox proportional hazards model was fitted to the data, and Schoenfeld residuals were extracted. The correlation between Schoenfeld residuals and ranked survival times for each covariate is presented in @tbl-table2.

```{r}
#| label: tbl-table2
#| tbl-cap: |
#|   [Schoenfeld Residual and Ranked Survival Time Correlation]{color="#FF2400"}
#| out.width: 40%
knitr::include_graphics(
	here::here("results", "tables", "t2.png")
)
```

While the p-values for the covariates triple negative subtype, grade, radiation, chemotherapy, and surgery are statistically significant at the 0.05 level, the observed correlations are not of large magnitude. Specifically, the correlations are -0.128, -0.042, 0.070, 0.110, and 0.074, respectively. Given that these values are relatively small in magnitude, we proceed with the Cox proportional hazards regression model.

\clearpage

@@

@tbl-table3 presents the median survival in years, along with 1-, 5-, and 10-year survival percentages, stratified by clinicopathological characteristics, all of which show statistically significant differences based on the log-rank test. Survival outcomes differ significantly by subtype, with the Triple-Negative subtype showing the poorest prognosis, including a median survival of 1.7 years and 1-, 5-, and 10-year survival rates of 67%, 24%, and 21%, respectively. Conversely, the Luminal B subtype has the most favorable outcomes, with a median survival of 9.4 years and a 10-year survival rate of 50%. Racial disparities are evident, as Black patients have the poorest survival outcomes, with a median survival of 2.4 years compared to 4.6 years for White patients and 5.4 years for patients of other races. Unmarried women experience poorer survival compared to married women. Although 1-year survival is similar across all node stages, nodal involvement significantly affects long-term outcomes, with the N3 stage showing the lowest median survival of 2.7 years and the poorest 5- and 10-year survival rates. Survival differences are also pronounced by distant metastasis (M stage), with M1 stage associated with worse outcomes. Additionally, higher cancer grades correlate with poorer survival. In terms of treatment, patients who receive radiation, chemotherapy, or surgery exhibit better survival outcomes across all measurements.

```{r}
#| label: tbl-table3
#| tbl-cap: |
#|   [Survival for IBC by Clinicopathological Characteristics]{color="#FF2400"}
#| out.width: 100%
knitr::include_graphics(
	here::here("results", "tables", "t3.png")
)
```

We conducted univariate and multivariate analyses based on the data in @tbl-table3, with the results of the Cox proportional hazards regression presented in @tbl-table4.

```{r}
#| label: tbl-table4
#| tbl-cap: |
#|   [Cox Proportional Hazards Regression
#|   Model for Survival in Adult Women with IBC]{color="#FF2400"}
#| out.width: 100%
knitr::include_graphics(
	here::here("results", "tables", "t4.png")
)
```

@@

In this large population-based cohort of women diagnosed with IBC, we found a better survival in Luminal B molecular subtype and poorer in Triple Negative. We found that Black patients made up the smallest percentage of Luminal B subtype and the greatest of Triple Negative. Similar findings were observed for patients who were unmarried at diagnosis.

Take things that ive noted from other study and compare results here.
It would be interesting to look at the combination of treatments (radiation, chemo, surgery)
Also comparing subtype to broader range of breat cancer not just IBC.

### 2024-12-03 products/manuscript/manuscript.tex
manuscript edits

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={epid 8010 manuscript},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{epid 8010 manuscript}
\author{}
\date{}

@@

\textbf{Authors}

\begin{itemize}
\tightlist
\item
  Murphy John\(^{1}\)
\end{itemize}

\textbf{Author affiliation}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Department of Epidemiology and Biostatistics, University of Georgia,
  Athens, GA, USA.
\end{enumerate}

@@

For each case, the following information from SEER was obtained: age at
diagnosis (20-49, 50-69, and \$\ge\$70), race (white, black, other),
marital status (unmarried, married), N stage (N0, N1, N2, N3), M stage
(M0, M1), grade (I/II, III/IV), radiation (no/unknown, yes),
chemotherapy (no/unknown, yes), surgery (no/unknown, yes), survival
months, cause-specific death classification (Alive or dead of other
cause, Dead (attributable to this cancer), Dead (missing/unknown COD)),
and other cause of death classification (Alive or dead due to cancer,
Dead (attributable to causes other than this cancer), Dead
(missing/unknown COD)). Breast cancer subtypes were classified as
Luminal A (HR+/HER2-), Luminal B (HR+/HER+), HER2 positive (HR-/HER2+),
and Triple Negative. In addition, cases with incomplete information on
any of these characteristics and patients who survived less than one
month were excluded from the study. Additionally, to focus on breast
cancer-specific mortality, patients with an unknown cause of death or
whose death was attributed to other causes were excluded.

@@

Clinicopatholical characteristics were compared between IBC subtype
using Pearson's Chi-squared test. Median, 1-year, 5-year, and 10-year
survival for IBC is estimated for each variable in the analysis using
the Kaplain-Meier method and variables were compared using the log-rank
test in the subgroups. Univariate and multivariate Cox proportional
hazards regressions were used to model the relationship between
potential covariates and IBC-specific survival. We obtained hazard
ratios (HR) and 95\% confidence intervals for each covariate. All
statistical analyses were performed using the survival package in R
version 4.3.3.

@@

A total of 5969 adult female IBC cases were eligible during 2010-2021 in
the SEER database for our study population. Of these, 1829 were
excluded; 165 had a survival time less than 1 month, 487 had an unknown
or competing cause of death, and 1177 had incomplete information. The
remaining 4140 were included in our analysis, as shown in Table 1. The
clinicopathological characteristics of the eligible patients are shown
in Table 2.

@@

@@

\section*{Discussion}\label{discussion}
\addcontentsline{toc}{section}{Discussion}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Impact of Molecular Subtype on Inflammatory Breast Cancer-Specific Mortality in Adult Women},
  pdfauthor={Murphy John},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Impact of Molecular Subtype on Inflammatory Breast
Cancer-Specific Mortality in Adult Women}
\author{Murphy John}
\date{}

@@

@@

For each case, the following information from SEER was obtained: age at
diagnosis (20-49, 50-69, and \textgreater=70), race (white, black,
other), marital status (unmarried, married), N (node) stage (N0, N1, N2,
N3), M (metastasis) stage (M0, M1), grade (I/II, III/IV), radiation
(no/unknown, yes), chemotherapy (no/unknown, yes), surgery (no/unknown,
yes), survival months, cause-specific death classification (Alive or
dead of other cause, Dead (attributable to this cancer), Dead
(missing/unknown COD)), and other cause of death classification (Alive
or dead due to cancer, Dead (attributable to causes other than this
cancer), Dead (missing/unknown COD)). Breast cancer subtypes were
classified as Luminal A (HR+/HER2-), Luminal B (HR+/HER+), HER2 positive
(HR-/HER2+), and Triple Negative. In addition, cases with incomplete
information on any of these characteristics and patients who survived
less than one month were excluded from the study. To focus on
IBC-specific mortality, patients with an unknown cause of death or whose
death was attributed to other causes were excluded.

@@

Clinicopathological characteristics were compared between IBC subtype
using Pearson's Chi-squared test. Median, 1-year, 5-year, and 10-year
survival for IBC is estimated for each variable in using the
Kaplain-Meier method and compared using the log-rank test in the
subgroups. The proportional hazards assumption was assessed visually
with log-log plots and statistically by calculating Schoenfeld residuals
and the Pearson's product moment correlation coefficient of the
Schoenfeld residuals and ranked survival times. After determining the
assumption of proportional hazard was met, univariate and multivariate
Cox proportional hazards regressions were used to model the relationship
between potential covariates and IBC-specific survival. We obtained
hazard ratios (HR) and 95\% confidence intervals for each covariate. All
statistical analyses were performed using the survival package in R
version 4.3.3.

\clearpage

@@

A total of 5969 adult female IBC cases were eligible during 2010-2021 in
the SEER database for our study population. Of these, 1829 were
excluded; 165 had a survival time less than 1 month, 487 had an unknown
or competing cause of death, and 1177 had incomplete information. The
remaining 4140 were included in our analysis, as shown in
Figure~\ref{fig-flowchart}.

\begin{figure}

\centering{

\includegraphics[width=0.9\textwidth,height=0.7\textheight]{../../results/figures/flowchart.png}

}

\caption{\label{fig-flowchart}{Flowchart of IBC patients screening on
SEER database}}

\end{figure}%

The clinicopathological characteristics of eligible patients are
summarized in Table~\ref{tbl-table1}. The age subgroup of 20--49 years
accounts for 35.1\% of Luminal B cases, a higher proportion compared to
other subtypes. Moreover, Luminal B and HER2-positive subtypes appear to
affect younger age groups more frequently than Luminal A, with 85.4\%
and 84.5\% of cases, respectively, occurring in patients under 69 years,
compared to 76.5\% for Luminal A. A higher proportion of Black patients
is observed in the Triple-Negative group compared to other subtypes.
Marital status analysis shows that the Triple-Negative group has the
highest percentage of unmarried women. Triple-Negative cases also have
the highest percentage of N3 node stage and exhibit the greatest
proportion of grade III/IV cases.

\begin{table}

\caption{\label{tbl-table1}{Clinicopathological characteristics by
cancer subtype}}

\centering{

\includegraphics{../../results/tables/t1.rds}

}

\end{table}%

\clearpage

\subsection{Proportional Hazards
Assumption}\label{proportional-hazards-assumption}

The proportional hazards assumption was assessed for each covariate
visually using log minus log plots, as shown in Figure~\ref{fig-loglog}.

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{../../results/figures/loglog_plot.png}

}

\caption{\label{fig-loglog}{Log minus log plots}}

\end{figure}%

Upon inspection, slight deviations from parallelism were observed for
the covariates subtype, grade radiation, chemotherapy, and surgery.
Therefore, the full Cox proportional hazards model was fitted to the
data, and Schoenfeld residuals were extracted. The correlation between
Schoenfeld residuals and ranked survival times for each covariate is
presented in Table~\ref{tbl-table2}.

\begin{table}

\caption{\label{tbl-table2}{Schoenfeld Residual and Ranked Survival Time
Correlation}}

\centering{

\includegraphics[width=0.4\textwidth,height=\textheight]{../../results/tables/t2.png}

}

\end{table}%

While the p-values for the covariates triple negative subtype, grade,
radiation, chemotherapy, and surgery are statistically significant at
the 0.05 level, the observed correlations are not of large magnitude.
Specifically, the correlations are -0.128, -0.042, 0.070, 0.110, and
0.074, respectively. Given that these values are relatively small in
magnitude, we proceed with the Cox proportional hazards regression
model.

\clearpage

@@

Table~\ref{tbl-table3} presents the median survival in years, along with
1-, 5-, and 10-year survival percentages, stratified by
clinicopathological characteristics, all of which show statistically
significant differences based on the log-rank test. Survival outcomes
differ significantly by subtype, with the Triple-Negative subtype
showing the poorest prognosis, including a median survival of 1.7 years
and 1-, 5-, and 10-year survival rates of 67\%, 24\%, and 21\%,
respectively. Conversely, the Luminal B subtype has the most favorable
outcomes, with a median survival of 9.4 years and a 10-year survival
rate of 50\%. Racial disparities are evident, as Black patients have the
poorest survival outcomes, with a median survival of 2.4 years compared
to 4.6 years for White patients and 5.4 years for patients of other
races. Unmarried women experience poorer survival compared to married
women. Although 1-year survival is similar across all node stages, nodal
involvement significantly affects long-term outcomes, with the N3 stage
showing the lowest median survival of 2.7 years and the poorest 5- and
10-year survival rates. Survival differences are also pronounced by
distant metastasis (M stage), with M1 stage associated with worse
outcomes. Additionally, higher cancer grades correlate with poorer
survival. In terms of treatment, patients who receive radiation,
chemotherapy, or surgery exhibit better survival outcomes across all
measurements.

\begin{table}

\caption{\label{tbl-table3}{Survival for IBC by Clinicopathological
Characteristics}}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{../../results/tables/t3.png}

}

\end{table}%

We conducted univariate and multivariate analyses based on the data in
Table~\ref{tbl-table3}, with the results of the Cox proportional hazards
regression presented in Table~\ref{tbl-table4}.

\begin{table}

\caption{\label{tbl-table4}{Cox Proportional Hazards Regression Model
for Survival in Adult Women with IBC}}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{../../results/tables/t4.png}

}

\end{table}%

@@

The multivariate analysis revealed that subtype, age, race, marital
status, N stage, M stage, grade, radiation, chemotherapy, and surgery
were independent predictors of IBC-specific survival. The Luminal B
subtype was associated with better survival (aHR=0.72, 95\%CI:
0.56-0.74, P\textless0.0001), whereas the Triple-Negative subtype showed
the poorest survival outcomes (aHR=2.51, 95\%CI: 1.79-2.21,
P\textless0.0001). In terms of treatment, our results find that the
existing therapies are effective in improving survival, that is
radiation (aHR=0.76, 95\%CI: 0.69-0.85, P\textless0.0001), chemotherapy
(aHR=0.50, 95\%CI: 0.44-0.57, P\textless0.0001), and surgery (aHR=0.59,
95\%CI: 0.52-0.66, P\textless0.0001).

\clearpage

\section{Discussion}\label{discussion}

In this large population-based cohort of women diagnosed with IBC, we
found a better survival in Luminal B molecular subtype and poorer in
Triple Negative. We found that Black patients made up the smallest
percentage of Luminal B subtype and the greatest of Triple Negative.
Similar findings were observed for patients who were unmarried at
diagnosis.

Take things that ive noted from other study and compare results here. It
would be interesting to look at the combination of treatments
(radiation, chemo, surgery) Also comparing subtype to broader range of
breat cancer not just IBC.
